Conversations in time: interactive visualisation to explore structured temporal data

### 2020-09-25 tsibbletalk.Rmd
minor corrections

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Department of Statistics
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Department of Econometrics and Business Statistics
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  Temporal data often has a hierarchical structure, defined by categorical variables describing different levels, such as political regions or sales products. Nesting of categorical variables produces a hierarchical structure. The \CRANpkg{tsibbletalk} package is developed to allow a user to interactively explore temporal data, relative to the nested or crossed structures. It can help to discover differences between category levels, and uncover interesting periodic or aperiodic slices. The package implements a shared `tsibble` object that allows for linked brushing between coordinated views, and a shiny module that aids in wrapping time lines for seasonal patterns. The tools are demonstrated using two data examples: domestic tourism in Australia and foot traffic in Melbourne.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

Temporal data typically arrives as a set of many observational units measured over time. Some variables may be categorical, indicating a hierarchy in the collection process, that may be measurements taken in different geographic regions, or types of products sold by the one company. Exploring these multiple features can be daunting. Ensemble graphics [@unwin_ensemble_2018] bundle multiple views of a data set together into one composite figure. These provide an effective approach for exploring and digesting many different aspects of temporal data. Adding interactivity to the ensemble can greatly enhance the exploration process.

This article describes new software, the \CRANpkg{tsibbletalk} package, for exploring temporal data using linked views and time wrapping. The next section provides some background to the approach based on setting up data structures and workflow. The section following explains the mechanism for constructing interactivity, to link between multiple hierarchical data objects and hence plots. The section after this describes the set up for interactively slicing and dicing time to wrap a series on itself to investigate periodicities.

@@

The \CRANpkg{tsibble} package [@wang2020tsibble] introduced a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. A `tsibble` extends the `data.frame` and \CRANpkg{tibble} classes with the temporal contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. An example of a `tsibble` can be found in the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package [@R-tsibbledata], shown below. The `Month` column holds year-months as the `index`. `State` and `Industry`, together, are the identifiers for these 152 series, which form the `key`. Note that the column `Series ID` could be an alternative option for setting up the `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent downstream temporal data analysis.

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data with many keys. The data has 152 series corresponding to different items in retail data. The multiple series are displayed using an overlaid time series plot, along with a scatterplot of two variables (trend versus seasonal strength) from feature space, where each series is represented by a dot. The feature space is computed using the `features()` function from \CRANpkg{feasts}, which summarises the original data for each series using various statistical features. This function along with other **tidyverts** functions is `tsibble`-aware, and outputs a table in a reduced form where each row corresponds to a series, which can be graphically displayed as in Figure \ref{fig:highlight-retail-2}.

@@

R \CRANpkg{shiny} [@R-shiny] and \CRANpkg{htmlwidgets} [@R-htmlwidgets] provides and infrastructure connecting R with HTML elements and JavaScript that provide the interactivity.<!--, although they seldom emphasise on interactive graphics.--> The \CRANpkg{htmlwidgets} package makes it possible to embed JavaScript libraries into R so that users are able to write only R code to generate web-based plots. Many JavaScript charting libraries have been ported to R as HTML widgets, including \CRANpkg{plotly} [@plotly2020], \CRANpkg{rbokeh} [@R-rbokeh], and \CRANpkg{leaflet} [@R-leaflet] for maps. Interactions between different widgets can be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk} [@R-crosstalk]. The \CRANpkg{crosstalk} extends \CRANpkg{htmlwidgets} with shared R6 instances to support linked brushing and filtering across widgets, without relying on \CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance built on a `tsibble`. This allows for seamless communication between different plots of temporal data. The `as_shared_tsibble()` function turns a `tsibble` into a shared instance,  `SharedTsibbleData`, which is a subclass of `SharedData` from \CRANpkg{crosstalk}). This is the object driving data transmission across multiple views. The \CRANpkg{tsibbletalk} package aims to streamline interactive exploration of temporal data, with the focus of temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical variable linking where selecting one or more observations in one category will broadcast to all other observations in this category. In addition, linking is also by time: within the time series plot, click on any data point, and  whole line will be highlighted in response. 
The `as_shared_tsibble()` uses `tsibble`'s `key` variables to achieve these types of linking.

The approach can also accommodate hierarchical temporal data. When one key variable is nested within another, such as regional areas within a state, this is considered to be a hierarchical structure. These time series are referred to as hierarchical and grouped time series in the literature [@fpp].  The `aus_retail` above is an example. Each series in the  data corresponds to all possible combinations of the `State` and `Industry` variables, which means they are intrinsically crossed with each other.

The `spec` argument in `as_shared_tsibble()`  constructing hybrid linking, that incorporates hierarchical and categorical linking. A symbolic formula can be passed to the `spec` argument, to incorporate structured specifications in the `key`. Adopting @Wilkinson1973's notation for factorial models, the `spec` follows the `/` and `*` operator conventions to declare nesting and crossing variables, respectively. The `spec` for the `aus_retail` data is therefore specified as `State * Industry` or `Industry * State`, which is the default for the presence of multiple `key` variables. If there is a hierarchy in the data, using `/` is required to indicate the parent-child relation, for a strictly one directional `parent/child`.

@@

In summary, shared the tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data using the common "key". The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linking between plots. The `plotly_key_tree()` function, in turn, decodes the hierarchical specification to plot a tree for data overview and navigation, when accompanied by more detailed plots.

@@

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and wrapping of the series into daily and weekly sections, respectively. Multiple seasonality pops out. There tends to be a daily pattern, especially visible at the main train station, Southern Cross. There is also a week day vs weekend pattern, also most visible at Southern Cross. These seasonal patterns are still present during the lockdown, but the magnitude is greatly reduced. Numbers are also down at the produce market and the shopping center. Birrarung Marr is the most affected. This is the location of special events and its clear that these have completely disappeared during lockdown.

The wrapping procedure involves slicing the time index into seasonal periods of interest, and the result is diced time. For example, hourly pedestrian data can be decomposed into 24-hour blocks, which then overlays the counts for all respective days, as done in plot \ref{fig:wrap-ped-2}. For exploration, this slice position should be controlled interactively, so that many different slices can be examined rapidly. This can be achieved using shiny, with the functions provided in a \CRANpkg{tsibbletalk}.

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a reusable code that could be embedded in any shiny application. In general, a shiny module provides a vehicle for modularising shiny applications, relevant for both users and developers. As with all shiny modules, the first argument in both functions in \CRANpkg{tsibbletalk} requires a user-supplied session id that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, so that the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. For date-times (represented by `POSIXt`), the granularity ranges from fine "day" to a much coarser "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

At the heart of the \CRANpkg{tsibbletalk} package is a blending the best bits from \CRANpkg{tsibble}, \CRANpkg{crosstalk}, \CRANpkg{plotly}, and \CRANpkg{shiny}.

>>>>

---
title: "Conversations in time: interactive visualisation to explore structured temporal data"
author:
- name: Earo Wang
  affiliation: The University of Auckland
  address: Department of Statistics
  email: earo.wang@auckland.ac.nz
- name: Dianne Cook
  affiliation: Monash University
  address: Department of Econometrics and Business Statistics
  email: dicook@monash.edu
bibliography: tsibbletalk.bib
abstract: >
  Temporal data often has a hierarchical structure, defined by categorical variables describing different levels, such as political regions or sales products. Nesting of categorical variables produces a hierarchical structure. The \CRANpkg{tsibbletalk} package is developed to allow a user to interactively explore temporal data, relative to the nested or crossed structures. It can help to discover differences between category levels, and uncover interesting periodic or aperiodic slices. The package implements a shared `tsibble` object that allows for linked brushing between coordinated views, and a shiny module that aids in wrapping time lines for seasonal patterns. The tools are demonstrated using two data examples: domestic tourism in Australia and pedestrian traffic in Melbourne.
preamble: |
  \usepackage{subfig}
output: rticles::rjournal_article
---

@@

Temporal data typically arrives as a set of many observational units measured over time. Some variables may be categorical, containing a hierarchy in the collection process, that may be measurements taken in different geographic regions, or types of products sold by the one company. Exploring these multiple features can be daunting. Ensemble graphics [@unwin_ensemble_2018] bundle multiple views of a data set together into one composite figure. These provide an effective approach for exploring and digesting many different aspects of temporal data. Adding interactivity to the ensemble can greatly enhance the exploration process.

This paper describes new software, the \CRANpkg{tsibbletalk} package, for exploring temporal data using linked views and time wrapping. We first provide some background to the approach based on setting up data structures and workflow, and give an overview of interactive systems in R. The section following introduces the \CRANpkg{tsibbletalk} package. We explain the mechanism for constructing interactivity, to link between multiple hierarchical data objects and hence plots, and describe the set up for interactively slicing and dicing time to wrap a series on itself to investigate periodicities.

@@

The \CRANpkg{tsibble} package [@wang2020tsibble] introduced a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. A `tsibble` extends the `data.frame` and `tibble` classes with the temporal contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. An example of a `tsibble` can be found in the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package [@R-tsibbledata], shown below. The `Month` column holds year-months as the `index`. `State` and `Industry` are the identifiers for these 152 series, which form the `key`. Note that the column `Series ID` could be an alternative option for setting up the `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent downstream temporal data analysis.

@@

Figure \ref{fig:highlight-retail} illustrates examining temporal data with many keys. The data has 152 series corresponding to different industries in retail data. The multiple series are displayed using an overlaid time series plot, along with a scatterplot of two variables (trend versus seasonal strength) from feature space, where each series is represented by a dot. The feature space is computed using the `features()` function from \CRANpkg{feasts}, which summarises the original data for each series using various statistical features. This function along with other **tidyverts** functions is `tsibble`-aware, and outputs a table in a reduced form where each row corresponds to a series, which can be graphically displayed as in Figure \ref{fig:highlight-retail-2}.

@@

R \CRANpkg{shiny} [@R-shiny] and \CRANpkg{htmlwidgets} [@R-htmlwidgets] provide infrastructure connecting R with HTML elements and JavaScript that support the interactivity. The \CRANpkg{htmlwidgets} package makes it possible to embed JavaScript libraries into R so that users are able to write only R code to generate web-based plots. Many JavaScript charting libraries have been ported to R as HTML widgets, including \CRANpkg{plotly} [@plotly2020], \CRANpkg{rbokeh} [@R-rbokeh], and \CRANpkg{leaflet} [@R-leaflet] for maps. Interactions between different widgets can be achieved with \CRANpkg{shiny} or \CRANpkg{crosstalk} [@R-crosstalk]. The \CRANpkg{crosstalk} extends \CRANpkg{htmlwidgets} with shared R6 instances to support linked brushing and filtering across widgets, without relying on \CRANpkg{shiny}.

@@

The \CRANpkg{tsibbletalk} package introduces a shared tsibble instance built on a `tsibble`. This allows for seamless communication between different plots of temporal data. The `as_shared_tsibble()` function turns a `tsibble` into a shared instance,  `SharedTsibbleData`, which is a subclass of `SharedData` from \CRANpkg{crosstalk}. This is the object driving data transmission across multiple views. The \CRANpkg{tsibbletalk} package aims to streamline interactive exploration of temporal data, with the focus of temporal elements and structured linking.

@@

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical variable linking where selecting one or more observations in one category will broadcast to all other observations in this category. That is, linking is by time: within the time series plot, click on any data point, and  whole line will be highlighted in response. 
The `as_shared_tsibble()` uses `tsibble`'s `key` variables to achieve these types of linking.

The approach can also accommodate temporal data of nesting and crossing structures. These time series are referred to as hierarchical and grouped time series in the literature [@fpp].  The `aus_retail` above is an example of grouped time series. Each series in the  data corresponds to all possible combinations of the `State` and `Industry` variables, which means they are intrinsically crossed with each other. When one key variable is nested within another, such as regional areas within a state, this is considered to be a hierarchical structure.

The `spec` argument in `as_shared_tsibble()` provides a means to construct hybrid linking, that incorporates hierarchical and categorical linking. A symbolic formula can be passed to the `spec` argument, to allow for structured specifications in the `key`. Adopting @Wilkinson1973's notation for factorial models, the `spec` follows the `/` and `*` operator conventions to declare nesting and crossing variables, respectively. The `spec` for the `aus_retail` data is therefore specified as `State * Industry` or `Industry * State`, which is the default for the presence of multiple `key` variables. If there is a hierarchy in the data, using `/` is required to indicate the parent-child relation, for a strictly one directional `parent/child`.

@@

In summary, shared tsibble data nicely bridges between the \CRANpkg{crosstalk} and **tidyverts** ecosystems for temporal data using the common "key". The `as_shared_tsibble()` provides a symbolic user interface for effortless construction of a hybrid of hierarchical and categorical linking between plots. The `plotly_key_tree()` function, in turn, decodes the hierarchical specification to plot a tree for data overview and navigation, when accompanied by more detailed plots.

@@

Figure \ref{fig:wrap-ped-2} and \ref{fig:wrap-ped-3} show slicing and wrapping of the series into daily and weekly sections, respectively. Multiple seasonalities pop out. There tends to be a daily pattern, especially visible at the main train station, Southern Cross Station. There is also a weekday vs weekend pattern, also most visible at Southern Cross Station. These seasonal patterns are still present during the lockdown, but the magnitude is greatly reduced. Numbers are also down at the produce market and the shopping center. Birrarung Marr is the most affected. This is the location of special events, and it is clear that these have completely disappeared during lockdown.

The wrapping procedure involves slicing the time index into seasonal periods of interest, and the result is diced time. For example, hourly pedestrian data can be decomposed into 24-hour blocks, which then overlays the counts for all respective days, as done in plot \ref{fig:wrap-ped-2}. For exploration, this slice position should be controlled interactively, so that many different slices can be examined rapidly. This can be achieved using shiny, with the functions provided in the \CRANpkg{tsibbletalk}.

This shiny module, decoupled to `tsibbleWrapUI()` and `tsibbleWrapServer()`, presents a clean interface and forms a reusable component that could be embedded in any shiny application. In general, a shiny module provides a vehicle for modularising shiny applications, relevant for both users and developers. As with all shiny modules, the first argument in both functions in \CRANpkg{tsibbletalk} requires a user-supplied session id that must be unique. The UI function `tsibbleWrapUI()` simply shows a slider that animates or controls the number of periods to be diced. The workhorse is the server function `tsibbleWrapServer()`, encapsulating the algorithm that transforms data and sends messages to update the plot accordingly. The `plot` argument expects a `ggplot` or `plotly` object, where one can plot data using either lines or other graphical elements (such as boxplots). As the function name suggests, a (shared) tsibble is needed to start the engine, so that the time `index` can be retrieved for dissection. The `period` option semantically takes a desired number of seasonal periods to be shifted, for example data shifted by "1 day", "2 days", or "1 week", etc. In other words, the `period` defines the grind level. For date-times (represented by `POSIXt`), the granularity ranges from fine "day" to a much coarser "year". The following code snippet generates Figure \ref{fig:wrap-ped}.

@@

At the heart of the \CRANpkg{tsibbletalk} package is a blending of the best bits from \CRANpkg{tsibble}, \CRANpkg{crosstalk}, \CRANpkg{plotly}, and \CRANpkg{shiny}.
